{\TheTitle}\thanks{Submitted to the editors DATE. \funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}

### 2018-07-03 ssc-0-0.tex
add a conclusion to the draft semi-implicit paper

% SIAM Article Template
\documentclass[review]{siamart1116}

@@

% REQUIRED
\begin{abstract}
The iterative semi-implicit formula, or some variant of it, is often
used to solve stochastic partial differential equations that arise
in quantum optics and in atom optics.  When this formula is applied
to a quantum system with unitary dynamics, stability depends on a
kind of Nyquist condition, where the angular frequency corresponding
to the largest eigenvalue of the discretised hamiltonian is at most
twice the reciprocal time step.  The formula develops parasitic
solutions very abruptly as this condition is exceeded, which makes
it troublesome for practical use, but quite useful as a test of the
stability of discretisation methods.
\end{abstract}

@@

% iterations
For a system of equations
\begin{equation}
	{ds\over dt}=f(s),
\end{equation}
the semi-implicit formula with time step~$h$ is
\begin{equation}\label{eq:i}
    s_{n+1}=s_n+hf_{n+\frac12},
	\qquad\text{where}\qquad 
    f_{n+\frac12}=f\left({\textstyle\frac{s_n+s_{n+1}}{2}}\right).
\end{equation}
It is not obvious how to implement this formula, because~$s_{n+1}$
and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
usual way is a simple iteration \cite{swx-5-12}.  To start, $s_{n+1}$~is initialised
to~$s_n$, and~$f_{n+\frac12}$ to~$f(s_n)$.  Then each iteration
evaluates Equation~\cref{eq:i}, and updates the
values of~$s_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the
iterations generate a sequence of~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
with
\begin{subequations}
\begin{align}
	\label{eq:iv} s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
	\label{eq:v} s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
		\qquad f_{n+\frac12}^{(j+1)}=f\left({\textstyle\frac{s_n+s_{n+1}^{(j)}}{2}}\right)\qquad \text{when $j\ge0$}.
\end{align}
\end{subequations}
This continues until the sequences converge, which is usually
achieved after a small, fixed number of iterations.  Any fixed point
of the iteration satisfies
Equation~\cref{eq:i}.  The remaining question is under what
conditions the iterations converge.

% steps stability
Generally, when a partial differential equation is discretised by
a given formula, the time step needs to be reduced in proportion
to the grid spacing in order for the discretised solution to remain
stable \cite{2000-Trefethen-Spectral}.  No one has analysed the stability of the
semi-implicit formula with iterations.  The stable time step has
had to be determined by trial and error, a time consuming process
for three-dimensional wave functions that are being simulated on
large grids.

This paper analyses the semi-implicit formula, in the case that it
is used to integrate linear Hamiltonian dynamics as in
Equation~\cref{eq:iii}.  The main result is that the stability
depends on a Nyquist-like condition, and the iterations converge
if and only if the eigenvalues of the hamiltonian are bounded by the reciprocal time step~$2/h$.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant of it, the laplacian operator is usually
evaluated by a spectral method \cite{2000-Trefethen-Spectral}.  This is equivalent to expanding
the wave function~$\psi(x)$ over a basis of sinc wave packets
\begin{equation}
	\psi(x) = \sum_m s^m\phi_m(x),
\end{equation}
where, for a spatial grid of step~$l$,
\begin{equation}
	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc x = \frac{\sin\pi x}{\pi x},
\end{equation}
under which the set of functions~$\phi_m$ is an orthonormal basis
for the space of functions limited to spatial bandwidth~$1/2l$.
The spectral method reduces Equation~\cref{eq:vi} to an ordinary
differential equation for the complex coefficient vector~$s$ of the
form
\begin{equation}\label{eq:iii}
	{ds\over dt}=-iHs,
\end{equation}
where~$H$ is a Hermitian matrix with elements
\begin{equation}
	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
\end{equation}

If the formula of Equation~\cref{eq:i} could be applied directly
to Equation~\cref{eq:iii}, the computed solution would be very
stable.  The coefficient vector~$s_n$ can be expanded over the
eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
projected onto the complex plane spanned by the eigenvector~$v_k$.
The exact solution that satisfies Equation~\cref{eq:iii}, with
initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
and its projection lies on the circle around the origin of
radius~$|c_k|$.  Here~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.

\begin{figure}[htbp]
  \centering
  \includegraphics{semii.1}
  \caption{A step of the semi-implicit formula, projected onto the
  complex plane spanned by the eigenvector~$v_k$ of~$H$.  The
  computed solution~$s_{n+1}$ lies on the same circle as the
  exact solution~$s(h)=\exp(-ihH)s_n$, but it rotates with angular frequency lower
  than~$\omega_k$.  See text for details.}
  \label{fig:iii}
\end{figure}

Suppose that the point~$s_{n+1}$ also lies on this circle.
Then the projection of~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
brought towards and away from the origin by moving~$s_{n+1}$ around
the circle, so there is some position for
which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, so that~$s_{n+1}$
satisfies Equation~\cref{eq:i}.  The only error introduced by discretisation
is that~$c_k$ rotates with an angular frequency less
than~$\omega_k$, bounded by~$\pi/h$ as~$\omega_k$ becomes
large.

In practice, the iterations of~Equation~\cref{eq:vi} must be used
instead of~Equation~\cref{eq:i}, and this is less stable.  For
Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
iteration rule becomes
\begin{equation}
s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
\end{equation}
or
\begin{equation}
s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
\end{equation}
with solution
\begin{equation}\label{eq:viii}
s_{n+1}^{(j)}=U^{(j)}s_n
	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
\end{equation}
The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
in~$-ih\omega_k/2$, which agrees to second order in~$h\omega_k$ with
the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$ is
plotted in \cref{fig:ii}, for the first four iterations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.  The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.  The exact time evolution operator~$U$ has unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The geometric series for~$U^{(j)}$ will converge if and only if the
angular frequencies~$\omega_k$ are bounded by the reciprocal time
step~$\frac2h$, a kind of Nyquist condition.  If the spectral
hamiltonian matrix has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
will enlarge the component of~$v_k$ in~$s_n$ by a factor close
to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
thus a parasitic solution growing at the
rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.

The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\nabla^2+V+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{1996-Trefethen-finite}, the mapping from~$\psi$ to $\partial\psi\over\partial t$ can be linearised at each time step, as
\begin{equation}
	\left.{\partial\psi\over\partial t}\right|_{\psi=\psi(t)+\delta\psi}
		\approx -i(\nabla^2+V+g|\psi(t)|^2)\psi(t) - i\hat H'(\psi(t))\delta\psi,
\end{equation}
where~$\hat H'(\psi(t))$ is an effective hamiltonian operator for
the field~$\delta\psi$.  This linearisation has been studied
extensively, and the dynamics generated by the hamiltonian~$\hat
H'$ are Bogoliubov sound waves \cite{aop-70-67}.  When this equation
is discretised with the semi-implicit formula, the stability condition
is that the Bogoliubov sound wave modes on a spatial grid of
spacing~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
fine grid, the high-frequency Bogoliubov modes have the same
frequencies as the eigenvectors of~$H$ \cite{2002-Pethick-Bose}.
Most simulations use grids fine enough for this to be the case, and
the nonlinear term will not affect their stability.

@@

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
  \label{fig:i}
\end{figure}

@@

@@

>>>>

% SIAM Article Template
% This needs to be compiled with LaTeX and dvipdf.  XeTeX doesn't work.
\documentclass[review]{siamart1116}

@@

% REQUIRED
\begin{abstract}
The iterative semi-implicit formula, or some variant of it, is often
used to solve stochastic partial differential equations that form part of Monte-Carlo methods used
in quantum optics and in atom optics.  When this formula is applied
to a quantum system with unitary dynamics, stability depends on a
kind of Nyquist condition, where the angular frequency corresponding
to the largest eigenvalue of the discretised hamiltonian is at most
twice the reciprocal time step.  The formula develops parasitic
solutions very abruptly as this condition is exceeded, which makes
it troublesome for practical use, but quite useful as a test of the
stability of discretisation methods.
\end{abstract}

@@

% iterations
For a system of ordinary differential~equations
\begin{equation}
	{ds\over dt}=f(s),
\end{equation}
the semi-implicit formula with time step~$h$ is
\begin{equation}\label{eq:i}
    s_{n+1}=s_n+hf_{n+\frac12},
	\qquad\text{where}\qquad 
    f_{n+\frac12}=f\left(\frac{s_n+s_{n+1}}{2}\right).
\end{equation}
It is not obvious how to implement this formula, because~$s_{n+1}$
and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
usual way is a simple iteration \cite{swx-5-12}.  To start, $s_{n+1}$~is initialised
to~$s_n$, and~$f_{n+\frac12}$ to~$f(s_n)$.  Then each iteration
evaluates Equation~\cref{eq:i}, and updates the
values of~$s_{n+1}$ and~$f_{n+\frac12}$.  This produces sequences~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
where
\begin{subequations}
\label{eq:v}
\begin{align}
	s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
	s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
		\qquad f_{n+\frac12}^{(j+1)}=f\left(\frac{s_n+s_{n+1}^{(j)}}{2}\right)\qquad \text{when $j\ge0$}.
\end{align}
\end{subequations}
This continues until the sequences converge, which is usually
achieved by a small, fixed number of iterations.  Any fixed point
of the iteration satisfies
Equation~\cref{eq:i}.  The question is under what
conditions the iterations converge.

Generally, when a partial differential equation is discretised by
a given formula, the time step needs to be reduced in proportion
to the grid spacing in order for the discretised solution to remain
stable \cite{2000-Trefethen-Spectral}.  No one has analysed the stability of the
semi-implicit formula with iterations.  The stable time step has
had to be determined by trial and error, a time consuming process
for three-dimensional wave functions that are being simulated on
large grids.

{\bf Make this less abrupt.}
This paper analyses the semi-implicit formula, in the case that it
is used to integrate linear Hamiltonian dynamics as in
Equation~\cref{eq:iii}.  The main result is that the stability
depends on a Nyquist-like condition, and the iterations converge
if and only if the eigenvalues of the Hamiltonian are bounded by the reciprocal time step~$2/h$.

@@

When the semi-implicit formula is applied to Equation~\cref{eq:vi}, or a
stochastic or nonlinear variant of it, the Laplacian operator is usually
evaluated by a spectral method \cite{2000-Trefethen-Spectral}.  This is equivalent to expanding
the wave function~$\psi(x)$ over a basis of sinc wave packets
\begin{equation}
	\psi(x) = \sum_m s^m\phi_m(x),
\end{equation}
where, for a spatial grid of step~$l$,
\begin{equation}
	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc x = \frac{\sin\pi x}{\pi x},
\end{equation}
under which the set of functions~$\phi_m$ is an orthonormal basis
for the space of functions limited to spatial bandwidth~$1/2l$.
The spectral method reduces Equation~\cref{eq:vi} to an ordinary
differential equation for the complex coefficient vector~$s$ of the
form
\begin{equation}\label{eq:iii}
	{ds\over dt}=-iHs,
\end{equation}
where~$H$ is a Hermitian matrix with elements
\begin{equation}
	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
\end{equation}
{\bf We're thinking about spectral collocation methods, with a delta function in place of~$\phi_m(x)$.  In the band-limited case, the collocation value is the same as the sinc component.  Explain this more fully.}

If the formula of Equation~\cref{eq:i} could be applied directly
to Equation~\cref{eq:iii}, the computed solution would be very
stable.  The coefficient vector~$s_n$ can be expanded over the
eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
projected onto the complex plane spanned by the eigenvector~$v_k$.
The exact solution to Equation~\cref{eq:iii}, with
initial condition~$s_n$ at time~$t_n$, is~$s(t_n+t)=\sum_k c^ke^{-i\omega_k(t_n+t)}v_k$,
where~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.   Its projection lies on the circle of
radius~$|c_k|$ centred on the origin.

\begin{figure}[htbp]
  \centering
  \includegraphics{semii.1}
  \caption{The numerical solution of Schr\"odinger's equation over
  a single time step~$h$, given by the ideal semi-implicit formula
  of Equation~\cref{eq:i}.  The trajectory of the discretised wave
  function~$s$ has been projected onto the complex plane spanned
  by the eigenvector~$v_k$ of the Hamiltonian operator.  The computed
  solution~$s_{n+1}$ follows the same circle as the exact
  solution~$s(t)=\exp(-iH(t-t_n))s_n$, but it rotates with a lower angular
  frequency than the eigenvalue~$\omega_k$.}
  \label{fig:iii}
\end{figure}

Suppose that the point~$s_{n+1}$ also lies on this circle, forming
a chord with~$s_n$.  In the projection shown, the point~$s_{n+\frac12}$
is the midpoint of the chord.  The
vector~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ points along the
chord.  Its magnitude is determined by how closely the chord passes
the origin, a distance that can be reduced at the same time
that~$|s_{n+1}-s_n|$ is increased by moving~$s_{n+1}$ further around
the circle.  There is always some position for
which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, satisfying Equation~\cref{eq:i}.
The only error introduced by discretisation is that~$c_k$ rotates
with an angular frequency less than~$\omega_k$, bounded by~$\pi/h$
as~$\omega_k$ becomes large.

In practice, the iterations of~Equation~\cref{eq:v} must be used
instead of~Equation~\cref{eq:i}, and this is less stable.  With the
assumption of Hamiltonian dynamics, the
iteration rule becomes
\begin{equation}
s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}\label{eq:x}
\end{equation}
or
\begin{equation}
s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
\end{equation}
with solution
\begin{equation}\label{eq:viii}
s_{n+1}^{(j)}=U^{(j)}s_n
	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
\end{equation}
The time step matrix~$U^{(j)}$ has the the same
eigenvectors~$v_k$ as the Hamiltonian matrix~$H$.  Each of its
eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
in~$-ih\omega_k/2$, which agrees to second order in~$h\omega_k$ with
the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$ is
plotted in \cref{fig:ii}, for the first four iterations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{wev.eps}
  \caption{The eigenvalue~$\omega_k$ of the Hamiltonian operator is compared to the size of the eigenvalue~$u_k^{(j)}$ of the time step operator~$U^{(j)}$ for $j$~iterations of the semi-implicit formula of Equation~\cref{eq:x}.  The exact time step operator has unimodular eigenvalues.  The lightest line shows~$j=1$ and the heavier ones show~$j=2$, 3, and 4.  The symbol~$h$ represents the time step, not Planck's constant.}
  \label{fig:ii}
\end{figure}

The geometric series for~$U^{(j)}$ will converge if and only if the
eigenvalues~$\omega_k$ are bounded by the reciprocal time
step~$\frac2h$, a kind of Nyquist condition.  If the spectral {\bf or discrete}
Hamiltonian matrix has a large eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
will grow the component of~$v_k$ in~$s_n$ by a factor close
to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component becomes a parasitic solution growing at the
rate
\begin{equation}
\lambda_k=\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right),\label{eq:xi}
\end{equation}
so the semi-implicit formula is unstable unless the~$\omega_k$ are bounded by~$\frac2{h}$.

The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\nabla^2+V+g|\psi|^2)\psi.
	label{eq:xii}
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{1996-Trefethen-finite}, the mapping from~$\psi$ to $\partial\psi\over\partial t$ can be linearised at each time step, as
\begin{equation}
	\left.{\partial\psi\over\partial t}\right|_{\psi=\psi(t)+\delta\psi}
		\approx -i(\nabla^2+V+g|\psi(t)|^2)\psi(t) - i\hat H'(\psi(t))\delta\psi,
\end{equation}
where~$\hat H'(\psi(t))$ is an effective Hamiltonian operator for
the field {\bf wave function?}~$\delta\psi$.  This linearisation has been studied
extensively: the dynamics generated by the Hamiltonian~$\hat
H'$ are Bogoliubov sound waves \cite{aop-70-67}.  The stability condition when this equation
is discretised with the semi-implicit formula
is that the Bogoliubov sound wave modes on a grid of
step~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
fine grid, the high-frequency Bogoliubov modes have the same
frequencies as the eigenvectors of~$H$ \cite{2002-Pethick-Bose}.
Most simulations use grids fine enough for this to be the case, and
the nonlinear term will not affect their stability.

@@

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{The growth of parasitic solutions when the dynamics of a quantum Kerr oscillator are computed with 4 iterations of the semi-implicit formula of Equation~\cref{eq:x}.  The coefficients~$s_n$ of the Hamiltonian eigenvectors should have constant sizes, shown by the parallel curves at~$t=0$ and~0.09.  The computed values are shown by the rulings between those lines.  Their sizes are constant for small~$n$, to the right of the dotted line at the stability limit~$\omega_n=2/h$.  Above the limit, the coefficients grow exponentially.  The growth rates predicted by Equation~\cref{eq:xi} are shown by the dashed line.}
  \label{fig:i}
\end{figure}

@@

\section{Discussion}

@@

For a wave function that follows Schr\"odinger's equation, the stability of the iterated semi-implicit formula is determined by the largest eigenvalue~$\omega$ of the Hamiltonian.  The formula is stable if~$\omega<\frac2{h}$, where~$h$ is the time step.  Otherwise, the eigenvector corresponding to~$\omega$ is a parasitic solution, whose rate of growth is stated in Equation~\cref{eq:xi} for the case~$\omega\gg\frac2{h}$, and can be read from from \cref{fig:ii} in general.

\Cref{fig:ii} suggests the performance of the formula when~$\omega\approx\frac2{h}$.  The discrete time step operator~$U^{(j)}$ differs significantly from the exact operator~$U$ for~$\omega>\frac1{h}$.  If the high-frequency dynamics are physically significant, a time step should be chosen well inside the limit of stability.

When the dynamics include a nonlinear term, as in Equation~\cref{eq:xii}, stability is determined by the largest eigenvalue of the linearised Hamiltonian about the solution.  In the common case of a Gross-Pitaevskii~equation, with grid spacing small compared to the healing length, the nonlinear term will have no effect on stability.

The stability limit is a kind of Nyquist condition, which makes it especially easy to reason about stability.  Physically, it is utterly reasonable for a simulation to require a time step on the characteristic time scale of the dynamics being simulated.
